---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# carbonr

<!-- badges: start -->
[![R-CMD-check](https://github.com/IDEMSInternational/carbonr/workflows/R-CMD-check/badge.svg)](https://github.com/IDEMSInternational/carbonr/actions)
[![Codecov test coverage](https://codecov.io/gh/IDEMSInternational/carbonr/branch/main/graph/badge.svg)](https://app.codecov.io/gh/IDEMSInternational/carbonr?branch=main)
[![Lifecycle:
experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![Project Status: WIP – Initial development is in progress, but there
has not yet been a stable, usable release suitable for the
public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![CRAN
status](https://www.r-pkg.org/badges/version/carbonr)](https://CRAN.R-project.org/package=carbonr)
[![license](https://img.shields.io/badge/license-LGPL%20(%3E=%203)-lightgrey.svg)](https://www.gnu.org/licenses/lgpl-3.0.en.html)
<!-- badges: end -->

 An R package to conveniently calculate carbon-equivalent emissions.
 
## Installation

You can install the development version of carbonr from [GitHub](https://github.com/) with:

``` r

# install.packages("devtools")
devtools::install_github("IDEMSInternational/carbonr")

```

## Example

```{r example}
library(carbonr)
# Want to calculate emissions for a flight between Vancouver and Toronto

airport_finder(name = "Vancouver")
airport_finder(name = "Toronto")
airplane_emissions("YVR","YYZ")

```

### Emissions calculated
                                                                                             
The emissions values used in the calculations are from the UK Government report (2021).

* airplane emissions
* ferry emissions
* hotel stay emissions
* office emissions
* rail emissions
* raw fuel emission data (primary source)
* vehicle emissions

                                                                                             
### Checking values alongside other calculators:
https://carbonfund.org/calculation-methods/
https://www.carbonfootprint.com/calculatorfaqs.html
                                                 
### Sources:
[1] UK government 2021 report. See https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/990675/2021-ghg-conversion-factors-methodology.pdf https://www.gov.uk/government/publications/greenhouse-gas-reporting-conversion-factors-2021
Note emissions for flights in the code uses values from direct effects only. Radiative forcing = TRUE will give indirect and direct effects. (multiples by 1.891).
See "business travel - air" sheet of gov.uk excel sheet linked above.

[2] Radiative forcing as 1.891 is from www.carbonfund.org
