---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# carbonr

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/carbonr)](https://CRAN.R-project.org/package=carbonr)
<!-- badges: end -->

 An R package to conveniently calculate carbon-equivalent emissions.
 
## Installation

You can install the development version of carbonr from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("IDEMSInternational/carbonr")```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(carbonr)
# Want to calculate emissions for a flight between Vancouver and Toronto

airport_finder(name = "Vancouver")
airport_finder(name = "Toronto")
airplane_emissions("YVR","YYZ")

```

## Emissions calculated
                                                                                             
The emissions that are calculated (with their respective sources given in square brackets):
                                                                                               
 * Train journeys[1]
 * Ferry/boat journeys[1]
 * Vehicles[1]
 * Flights[1, 2]
 * Secondary Emissions[3]
 * Hotel stays[1]
 
Sources for the emission values
                                                                                             
[1] UK government 2021 report. See https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/990675/2021-ghg-conversion-factors-methodology.pdf https://www.gov.uk/government/publications/greenhouse-gas-reporting-conversion-factors-2021
Note emissions for flights in the code uses values from direct effects only. Radiative forcing = TRUE will give indirect and direct effects. (multiples by 1.891).
See "business travel - air" sheet of gov.uk excel sheet linked above.

[2] Radiative forcing as 1.891 is from www.carbonfund.org

[3] Secondary sources calculated using https://www.carbonfootprint.com/calculatorfaqs.html

## Next Steps:
 * checkmate package for errors
 * Link up with Shiny app calculator?
 * Create simple vignette
 * Add in short haul/medium haul for flights - Under "Employee Travel" https://carbonfund.org/calculation-methods/
"Short flights are calculated to be under 300 miles one-way with emissions of 0.217kg CO2e per passenger mile
Medium flights are calculated to be 300-2300 miles one-way, average 1500 miles, with emissions of 0.134 kg CO2e per passenger mile
Long flights are calculated to be > 2300 miles, average 3,000 miles one-way with emissions of 0.167kg CO2e per passenger mile"
                                                                                             
                                                                                             
## Checking values alongside other calculators:
https://carbonfund.org/calculation-methods/
https://www.carbonfootprint.com/calculatorfaqs.html
                                                                                             
