---
title: "Company Emissions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Company Emissions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Company Emissions with carbonr
 This vignette demonstrates a step-by-step process for calculating and combining different types of company emissions, including office emissions, hotel emissions, and travel emissions, using various functions to track and calculate the carbon footprint associated with different company activities.


### Introduction
Understanding the overall carbon footprint of a company is crucial for meeting sustainability goals. This vignette details the process for calculating emissions from company office operations, employee travel (air, ferry, rail), and hotel accommodations, ultimately allowing the organization to identify high-emission areas and take action to reduce them.

To begin, install and load the necessary packages.
```{r install_and_loading_packages, eval = FALSE}
install.packages("emissionsPkg")
library(emissionsPkg)
install.packages("carbonr")
library(carbonr)
library(tidyverse)
```
## Office Emissions:
Start by calculating emissions from office-related activities such as energy consumption, waste production, and equipment usage.

### Example Office Emission Calculation
 For calculating office emissions, use `office_emissions()` function from the `carbonr` package. This function requires input values for energy use (e.g., electricity and heating).

```{r office_emissions_example, message=FALSE, warning=FALSE}
 office_emissions <- carbonr::office_emissions(specify = TRUE,
   electricity_kWh = 200,
   water_supply = 100,
   water_unit = "cubic metres",water_trt = FALSE
 )
 knitr::kable(office_emissions)
```

### Hotel Emissions:
 Next, we calculate emissions related to hotel stays for business trips using `hotel_emissions()`. This function calculates the emissions based on the number of nights, room type, and occupancy rate.

### Example Hotel Emission Calculation
```{r hotel_emissions_example, message=FALSE, warning=FALSE}
hotel_emissions <- carbonr::hotel_emissions(location = "Fiji", nights = 2, rooms=1)
 knitr::kable(hotel_emissions)
```
### Travel-related Emissions:

## 1. Airplane Emissions
Using the `carbonr` package, we can calculate emissions for a flight between Nairobi and Kisumu by first finding the airport codes.
### Finding Airport Codes
```{r find_airport_nairobi, eval=FALSE, message=FALSE, warning=FALSE}
 airport_finder(name = "Nairobi") %>% knitr::kable()
```
```{r find_airport_kisumu, eval=FALSE, message=FALSE, warning=FALSE}
 airport_finder(name = "Kisumu") %>% knitr::kable()
```

### Calculating Flight Emissions
With the IATA codes (`"NBO"` for Nairobi and `"KIS"` for Kisumu), we can calculate emissions:
```{r airplane_emissions_chunk, message=FALSE, warning=FALSE}
result <- carbonr::airplane_emissions("NBO", "KIS")  
knitr::kable(result)
```

## 2. Ferry Emissions
To estimate emissions for ferry travel, for instance, from Shimoni to Giohar, we start by locating seaport codes.

### Finding Seaport Codes
```{r shimoni_seaport, eval=FALSE, message=FALSE, warning=FALSE}
seaport_finder(city = "Shimoni") %>% knitr::kable()
```

```{r giohar_seaport, eval=FALSE, message=FALSE, warning=FALSE}
seaport_finder(city = "Giohar") %>% knitr::kable()
```

## Ensure the seaports data is loaded
```{r load_seaports, message = FALSE, warning = FALSE}
data("seaports", package = "carbonr")
```

### Calculating Ferry Emissions
With codes `"SMN"` for Shimoni and `"GHR"` for Giohar:
```{r ferry_example, message=FALSE, warning=FALSE}
result <- carbonr::ferry_emissions("SMN", "GHR", round_trip = TRUE)
knitr::kable(result)
```
## 3. Train Emissions
Train journey emissions are calculated similarly. To estimate emissions from Bristol Temple Meads to Edinburgh Waverley via Birmingham New Street:

### Finding Station Names
```{r load_stations, message=FALSE, warning=FALSE}
data("stations", package = "carbonr")
```

```{r find_station_bristol, eval=FALSE, message=FALSE}
rail_finder("Bristol") %>% knitr::kable()
```

```{r find_station_edinburgh, eval=FALSE, message=FALSE}
 rail_finder("Edinburgh") %>% knitr::kable()
```

### Calculating Train Emissions
```{r rail_em, message=FALSE, warning=FALSE}
result <- carbonr::rail_emissions(from = "Bristol Temple Meads", to = "Edinburgh", via = "Birmingham New Street")
knitr::kable(result)
```

## 4. Aggregating Emissions for Multiple Individuals
Using data frames and mapping functions, we can calculate combined emissions for multiple travel types.

### Example
```{r, eval=FALSE}
multiple_ind <- tibble::tribble(
 ~ID, ~rail_from, ~rail_to, ~air_from, ~air_to, ~air_via,
"Clint", "Bristol Temple Meads", "Paddington", "LHR", "KIS", "NBO",
"Zara", "Bristol Temple Meads", "Paddington", "LHR", "LAX", "ORL"
 )

multiple_ind %>%
rowwise() %>%
mutate(
plane_emissions = carbonr::airplane_emissions(air_from, air_to, air_via),
train_emissions = carbonr::rail_emissions(rail_from, rail_to)
) %>%
mutate(total_emissions = plane_emissions + train_emissions) %>%
knitr::kable()
```

## References:
Emission factors are based on the UK Government’s 2022 greenhouse gas reporting conversion factors: [Greenhouse Gas Conversion Factors](https://www.gov.uk/government/publications/greenhouse-gas-reporting-conversion-factors-2022).
