## Overview

carbonr is a package in R to conveniently calculate carbon-equivalent emissions.

The emissions values in the calculations are mostly from the [UK Government report (2023)](https://www.gov.uk/government/publications/greenhouse-gas-reporting-conversion-factors-2023) when available.


## Motivation
`carbonr` aims to provide a reliable and reproducible approach to calculating emissions levels, ensuring that the results can be saved, edited, and redistributed easily. Further, `carbonr` aims to be transparent in its calculations and values applied. This has the bonus of allowing for flexibility and customisation in estimating carbon-equivalent emissions.

The vision for `carbonr` is to expand and become more comprehensive. We invite contributions from the community to extend the package's functionality, build additional features, and transform it into a more robust tool for estimating carbon-equivalent emissions. Similarly, we invite open discussions and contribution to capture different perspectives and enhancing the functionality of the package.

Finally, `carbonr` aims to make the estimation of carbon-equivalent emissions more accessible by offering a user-friendly front-end interface using Shiny. This ensures that the tools are easier to use, even for individuals with limited programming experience.

## Installation

You can install the development version of carbonr from [GitHub](https://github.com/):

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("IDEMSInternational/carbonr")
```

## Usage

The main function in `naflex` is `na_omit_if`. 

When wrapped around a vector in a summary function, `na_omit_if` ensures that the summary value is calculated when the checks pass, and returns `NA` if not. The example below shows how to calculate the `mean`, conditionally on the proportion of missing values.

```{r}
library(naflex)

x <- c(1, 3, NA, NA, 3, 2, NA, 5, 8, 7)

# Calculate if 30% or less missing values
mean(na_omit_if(x, prop = 0.3))
# Calculate if 20% or less missing values
mean(na_omit_if(x, prop = 0.2))
```

Four types of checks are available:

1. `prop`: the maximum proportion (0 to 1) of missing values allowed
2. `n`: the maximum number of missing values allowed
3. `consec`: the maximum number of consecutive missing values allowed, and
4. `n_non`: the minimum number of non-missing values required.

If multiple checks are specified, all checks must pass for missing values to be removed. For example, although there are less than 4 missing values in `x`, there are two consecutive missing values, hence the `consec = 1` check fails in the example below the result is `NA`.

```{r}
# Calculate if 4 or less missing values and 1 or less consecutive missing values 
mean(na_omit_if(x, n = 4, consec = 1))
```
The use of `%>%` ("pipe") from `magrittr` can be used to make the code look clearer and more familiar. The beginning of the line is now the same as standard R and it moves `na_omit_if` after `x` which then appears more like an option within the function, like `na.rm`, which is how you might think about `na_omit_if` conceptually in this case.

```{r, warning = FALSE}
require(magrittr)

sum(x %>% na_omit_if(prop = 0.25))
```

Note that you should not use `na_omit_if` with `na.rm = TRUE` in the summary function since this will always remove missing values so the checks are essentially ignored.

## How `naflex` works & more details

`na_omit_if` works by removing the missing values from `x` if the checks pass, and leaving `x` unmodified otherwise.

```{r}
# Missing values removed
na_omit_if(x, n = 4)
```
`na_omit_if` can be thought of like an extension of `stats::na.omit` and if missing values are removed, an `na.action` attribute and `omit` class are added for consistency with `stats::na.omit`.

```{r}
# Missing values not removed, x is unmodified
na_omit_if(x, n = 2)
```
A further set of four `na_omit_if_*` functions are provided for doing the same thing but restricted to a single check e.g. `na_omit_if_n(x, 2)`.

`na_check` has the same parameters as `na_omit_if` but returns a logical indicating whether the checks pass. It is used internally in `na_omit_if` and may also be a useful helper function. 

```{r}
if (na_check(x, n = 4, consec = 1)) "NA checks pass" else "NA checks fail"
```

A set a four `na_check_*` functions are also provided for doing the same thing restricted to a single check e.g. `na_check_prop(x, 0.2)`

Finally, `naflex` provides a set of helper functions for calculating missing value properties used in these checks.

```{r}
na_prop(x)
na_n(x)
na_consec(x)
na_non_na(x)
```

## Compared to base R

In base R, this functionality can often be achieved using a combination of `ifelse`, `is.na`, `rle` and the option `na.rm = TRUE`.`naflex` aims to simplify, shorten and standardise this process for users.

For example, the equivalent of:

```{r}
mean(na_omit_if(x, n = 4, prop = 0.2))
```

in base R is:

```{r}
ifelse(sum(is.na(x)) <= 4 && mean(is.na(x)) <= 0.2, mean(x, na.rm = TRUE), NA)
```

The check for longest sequence of consecutive missing values is more complex and requires clever use of the `rle` function. For example,

```{r}
mean(na_omit_if(x, consec = 5))
```

is equivalent to:

```{r}
r <- rle(is.na(x))
m <- r$lengths[r$values]
ifelse(max(m) <= 5, mean(x, na.rm = TRUE), NA)
```


## Theatre Waste Functions
There are additional more specific functions related to operating theatre waste, alternative sources are used given in the References section.

Carbon credit prices are additionally available in the `carbon_credit_price` function where values are based on the [World Bank data](https://carbonpricingdashboard.worldbank.org/). The jurisdiction and year available for that jurisdiction can be found in the `check_CPI()` function.

## References
<sup id="f1">1</sup> <a href="https://library.wmo.int/index.php?lvl=notice_display&id=20130#.XljKS84zZnI" target="_blank">WMO Guidelines on the Calculation of Climate Normals</a> [↩](#a1)